### 2014-09-19 text/Chapter06.tex
Start working on Chapter 4
Calculus and semantics work (Chapter 5)
Cleanup files, fix some invalid references

% ==================================================================================================
% Macros - for ICFP 2014
% ==================================================================================================

% comonadic semantics
\newcommand{\F}{\mathsf{F}}
\newcommand{\D}{\mathsf{D}}
\newcommand{\Ds}[3]{\D^{\sclrd{#1}}_{#2} #3}
\newcommand{\Icat}{\cclrd{\mathbb{I}}}
\newcommand{\id}{\mathit{id}}
\newcommand{\B}[1]{\bar{#1}}
\newcommand{\bbullet}{\mathop{\bar{\bullet}}}
\renewcommand{\idf}[0]{ \ident{id} }
\newcommand{\m}{\mathsf{m}}
\newcommand{\n}{\mathsf{n}}

\newcommand{\shortar}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaro}[4]{\ar@{}[#3]^(.1){}="#1"^(.8){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\shortarn}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar_{#4}"#1";"#2"}
\newcommand{\shortaron}[6]{\ar@{}[#3]^(#5){}="#1"^(#6){}="#2"\ar^{#4}"#1";"#2"}
\newcommand{\diagar}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}^{#4}"#1";"#2" }
\newcommand{\diagaro}[4]{\ar@{}[#3]^(.15){}="#1"^(.85){}="#2"\ar@{<=}_{#4}"#1";"#2" }

% Typing rule and typing statement (single line)
\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
% Context transformation and context transformation with new line in the middle
\renewcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}

\newcommand{\czero}{ \textcolor{cclr}{ \mathsf{ign} } }
\newcommand{\cunit}{ \textcolor{cclr}{ \mathsf{use} } }
\newcommand{\cseq}{ \textcolor{cclr}{ \circledast }} % \bullet} } % \textcolor{cclr}{\bullet}}  
\newcommand{\cpar}{ \textcolor{cclr}{\oplus} }
\newcommand{\cleq}{ \textcolor{cclr}{\leq} }

% \newcommand{\single}[{\langle{#1}\rangle}

\newcommand{\uncurry}{\Lambda^{\hspace{-0.2em}-1}}

% Coeffect shapes

\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\renewcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }

\newcommand{\sflat}{\star}

\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
\newcommand{\aseq}{ \textcolor{aclr}{\circledast}} % Should be the same as \cseq 
% Tomas: I prefer \circledast here because it is a circled (lifted) version of \cseq 
% { \textcolor{aclr}{\bullet}}
\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }

% Free monoid used in the *structural* coeffect algebra
\newcommand{\fcons}{ \times } 
\newcommand{\fempty}{\epsilon}

@@

\emph{Context} is important for defining meaning-- not just in natural languages, but also in logics and 
programming languages.
The standard notion of context in programming is an environment 
providing values for free variables. An open term with free variables is context dependent -- its 
meaning depends on the free-variable context. The simply-typed $\lambda$-calculus famously 
analyses such context usage. 
Other systems go further. %capturing how 
%the context is used. 
For example, bounded linear logic tracks the number of times a 
variable is used~\cite{girard1992bounded}.

In software engineering, ``context'' provides more than just free-variable values. For example, in a
distributed system, the context provides different resources that may be available on different 
devices (\eg{}, a database on a server or a GPS sensor on a phone).

In this paper we develop a calculus for capturing various notions of 
context in programming. A key feature and contribution
of the calculus is its \emph{coeffect system} which provides a static analysis for contextual properties (coeffects). The system
follows the style of type and effect systems, but captures a different class of properties. % (Section~\ref{sec:intro-pathways}). 
Another key contribution of the calculus is its semantics which can be smoothly instantiated for specific notions of context.
%The semantics of the calculus is general and can be instantiated for particular notions of context.

Coeffect systems were previously introduced as a generic analysis of
context dependence which can be instantiated for various notions of
context~\cite{petricek2013coeffects}. However, the formalization was restricted to
tracking a class of \emph{whole-context} properties where a term has just
one coeffect. This limited the applications and precision of any
analysis. For example, a whole-context liveness analysis marks the 
free-variable context as live (some variable may be used) or dead 
(no variable is used), but it cannot record liveness \emph{per-variable}.

We develop a more general system which captures both
\emph{per-variable} coeffects, which we call \emph{structural}, and
\emph{whole-context} coeffects, which we call \emph{flat}, and more. Our
key contributions are:

\begin{itemize}
\item We present a general coeffect type system (Section~\ref{sec:coeffects}) and demonstrate two 
  concrete uses -- \emph{flat} coeffect systems that tracks whole-context information and 
  \emph{structural} coeffect systems which tracks fine-grained per-variable information.

\item We show practical examples, instantiating the calculus for structural systems capturing 
  variable usage based on bounded linear logic, dataflow caching, and precise liveness analysis.
  We also instantiate the calculus to flat systems, building on and extending previous examples~\cite{petricek2013coeffects}.
% variants of the coeffect system capture examples described   in our earlier work

\item We discuss the syntactic properties of the \emph{structural} coeffect system 
  (Section~\ref{sec:syntax}). It satisfies type preservation under both $\beta$-reduction 
  and $\eta$-expansion, allowing its use with both call-by-name and call-by-value languages.
  This important property distinguishes it from both effect systems and flat coeffects.

\item We provide a denotational semantics, resisting and extending the notion of \emph{indexed comonads} to the 
structural setting (Section~\ref{sec:semantics}). We prove soundness by showing 
  the correspondence between syntactic and semantic properties of coeffect systems.
\end{itemize}

\noindent
Coeffects can be approached from multiple directions (Section~\ref{sec:intro-pathways}) including 
syntactic (effect systems), semantic, and proof-theoretic. We 
emphasize the syntactic view, though we also outline a categorical semantics
and note the interesting technical details.

@@

\section{The coeffect calculus}
\label{sec:coeffects}

The three calculi shown in the previous section track two kinds of
contextual properties: bounded reuse and dataflow are structural
(per-variable) systems, and implicit parameters and our earlier coeffect
systems~\cite{petricek2013coeffects} are flat (whole-context) systems. 
This section presents our primary contribution: the general coeffect
calculus.

The calculus is parameterised by an algebraic structure of coeffects.
To capture both structural and flat systems, coeffect annotations are
indexed by a \emph{shape}. In flat systems, the shape is a singleton set
and so annotations are \emph{scalar} values. Structural systems use shapes
matching the number of variables in a free-variable context and so annotations
are \emph{vectors}. However, the coeffect calculus could also use shapes describing
trees and other structures.

\subsection{Understanding coeffects: syntax and semantics}

The coeffect calculus provides both an analysis of
context dependence (its coeffect system) and a semantics for
context (see Section~\ref{sec:semantics}). These two features of the calculus provide different
perspectives on coeffect annotations $\aclrd{R}$ in a judgment $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$:
%%
\begin{itemize}
\item Syntactically, coeffects model \emph{contextual requirements} and 
  may be overapproximated, such that more capabilities are required 
  than necessary at runtime.

\item Semantically, coeffects model \emph{contextual capabilities} and behave 
  like containers of capabilities, such that the semantics may throw away 
  capabilities that will not be needed.
\end{itemize}
%
Thus there are two dual ways to understand coeffect annotations. Each
perspective implies an alternate reading of the typing rules.
%
\begin{itemize}
\item
As \emph{contextual requirements}, the rules should be read top-down.
The requirements of multiple sub-terms are \emph{merged} and the
requirements of function body are \emph{split} between immediate (declaration-site)
and latent (call-site) coeffects.

\item
As \emph{contextual capabilities}, the rules should be read bottom-up.
The capabilities provided to a larger term are \emph{split} between 
sub-terms; for functions, the capabilities of declaration-site and call-site
are \emph{merged} and passed to the body.
\end{itemize}
%
The reason for this asymmetry follows from the fact that context appears in a \emph{negative
position} in the model. In Section~\ref{sec:semantics}, the denotation of a judgment
is a function of the form $\D_{\aclrd{R}} \interp{\Gamma} \rightarrow \interp{\tau}$ where $\D_{\aclrd{R}} \interp{\Gamma}$ 
encodes the contextual capabilities used to evaluate a term. Similarly
functions have models of the form $\D_{\cclrd{s}} \interp{\sigma} \rightarrow \interp{\tau}$
with additional contextual capabilities attached to the input.

@@

\noindent
The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
category in the semantics (Section~\ref{sec:semantics}). 
It models sequential composition with variable access
($\cunit$) as the identity. The other element ($\czero$) is used for
variables that are not accessed. The operation $\cpar$ combines coeffects
for contexts used in multiple places (contraction). 
The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
not always mean $\ast$ and $+$.

The context annotations can be viewed as \emph{containers} of scalar coeffects.
For structural coeffects, the container is a vector, while for flat coeffects, it is
a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{abbott2005containers}
which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.

@@

\newcommand{\narrow}[1]{\hspace{-0.6em}#1\hspace{-0.6em}}

@@

\[
\tystmt{weak}
  { \quad\quad 
    \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x:\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}
    \quad\quad\quad }
\]
\[ 
\tystmt{exch}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau, y:\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
               { \quad \coctx{\Gamma_1, y:\sigma, x:\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }{\emptyset} }
\]
\[
\tystmt{contr}
  { \ctxtransnl{ \coctx{\Gamma_1, y:\tau, z:\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]}}
\]
\[
\tystmt{sub}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau,\Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \aclrd{T} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s'}} \,\asplit\, \aclrd{T}} }{\emptyset} }
\quad(\cclrd{s} \;\cleq\; \cclrd{s'} )                 
\]
\caption{The general coeffect calculus}
\vspace{-1em}
\label{fig:coeffects-general}
\end{figure}

@@

\subsection{Structural coeffects}
\label{sec:coeffects-structural}

@@

\subsection{Flat coeffects}
\label{sec:coeffects-flat}

@@

\section{Equational theory}
\label{sec:syntax}
\newcommand{\reduce}{\leadsto}

Each of the concrete coeffect systems discussed in this paper has a different notion of 
context-dependence, much like various effectful languages have different notions of effects
(such as state or exceptions). However, there are equational properties that hold for all (or some) of
the systems we consider.

The equational theory in this section illuminates 
the axioms of coeffect algebra and the semantics of the calculus.
We discuss syntactic substitution as it can form the basis for
reduction in a concrete operational semantics. 
We consider structural and flat systems
separately. This provides better insight into how the two systems work and
differ. In particular, call-by-name evaluation is \emph{coeffect
preserving} for all structural, but only some flat systems.

The properties and proofs in this section are syntactic.
In Section~\ref{sec:semantics-soundness} we show that
our denotational model of the coeffect calculus is sound with respect
to the equational theory here.

We use standard syntactic substitution written as $\subst{e_1}{x}{e_2}$, 
$\beta\textnormal{-reduction}$ and $\eta$-expansion, written as $\reduce_\beta$ and $\reduce_\eta$.
Equality of terms $e_1$ and $e_2$, defined by a relation
$\equiv$, requires the equality of their contexts, types, and 
coeffects, written $\coctx{\Gamma}{\aclrd{R}} \vdash e_1 \equiv e_2 : \tau$.

%\begin{definition}[Subject reduction]
%Given reduction $e \reduce e'$ and $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$
%then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$, \ie{}, $\reduce$ is 
%\emph{type-and-coeffect preserving}. 
%\label{def:pres}
%\end{definition}

%Definition~\eqref{def:pres} is also known as the \emph{subject reduction} property.

\subsection{Structural coeffect systems}
\label{prop:structural}

For structural coeffect systems, recall that coeffects are vectors
with $\asplit = \amerge = \cons$ (vector concatenation) and $\azero =
\langle{}\rangle$ (the empty vector), thus coeffect annotations
comprise the \emph{free monoid} \ie, lists over the coeffect scalars
(although we continue using the vector terminology). We first show
substitution:

\begin{lemma}[Substitution lemma] 
\label{thm:structural-subst}
In a structural coeffect calculus with a coeffect scalar structure $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$: 
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{S}} \vdash e_s : \sigma  
\;\;  \wedge  \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{\aclrd{R_1} \cons\; \alift{\cclrd{r}} \cons\; \aclrd{R_2} } \vdash e_r : \tau\\
\quad \Rightarrow \; \coctx{\Gamma_1,\Gamma,\Gamma_2}{\aclrd{R_1} \cons\; (\cclrd{r} \,\aseq\, \aclrd{S}) \cons\; \aclrd{R_2}} \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\label{lemm:structural-subst}
\end{lemma}
\begin{proof}
By induction over the $e_2$ derivation using the free monoid structure ($\C, \cons, \langle{}\rangle$)
and coeffect scalar axioms (full proof~\cite{appendix}). %Full proof in the Appendix.
\end{proof}

\noindent
Because of the vector (free monoid) structure, coeffects $\aclrd{R_1}$, $\aclrd{R_2}$, and 
$\langle{\cclrd{r}}\rangle$ for the receiving term $e_r$ are uniquely associated with $\Gamma_1$, $\Gamma_2$, and
$x$ respectively. Therefore, substituting $e_s$ (which has coeffects $S$) for $x$ introduces
the context dependencies specified by $S$ which are composed with the requirements $r$ on $x$. 
Using the substitution lemma, we can demonstrate $\beta$-equality:
%
\begin{equation*}
\inference{
  \inference{\coctx{\Gamma_1, x : \sigma}{\aclrd{R} \cons \langle{\cclrd{r}}\rangle} \vdash e_1 : \tau}
             {\coctx{\Gamma_1}{\aclrd{R}} \vdash \lambda x . e_1 : \sigma \xrightarrow{\cclrd{r}} \tau} &
           \coctx{\Gamma_2}{\aclrd{S}} \vdash e_2 : \sigma}
  {\coctx{\Gamma_1,\Gamma_2}{\aclrd{R} \cons (\cclrd{r} \,\aseq\, \aclrd{S})} \vdash (\lambda x . e_1) e_2 
\equiv
\subst{e_1}{x}{e_2} : \tau}
\end{equation*}
%%
\noindent
As a result, $\beta$-reduction preserves the type and coeffects of a term.
This gives the following subject reduction property:

\begin{theorem}[Subject reduction]
In a structural coeffect calculus, if $\coctx{\Gamma}{\aclrd{R}} \vdash e : \tau$ and
$e \reduce_\beta e'$ then $\coctx{\Gamma}{\aclrd{R}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
Following from Lemma~\ref{thm:structural-subst} and $\beta$-equality.  
\end{proof}

\noindent
Structural coeffect systems also exhibit $\eta$-equality, therefore satisfying both
\emph{local soundness} and \emph{local completeness} conditions set by Pfenning and Davies
\cite{logic-modal-reconstruction}. This means that abstraction does not introduce too much, 
and application does not eliminate too much.
%
\begin{equation*}
\inference{
  \inference 
    {\coctx{\Gamma}{\aclrd{R}} \vdash e : \sigma \xrightarrow{s} \tau & 
     \coctx{x:\sigma}{\alift{\cunit}} \vdash x : \sigma}
    {\coctx{\Gamma, x:\sigma}{\aclrd{R} \,\cons\, (\cclrd{s} \,\aseq\, \alift{\cunit})} \vdash e \, x : \tau}}
  {\coctx{\Gamma}{\aclrd{R}} \vdash \lambda x.e~x  \equiv e 
: \sigma \xrightarrow{\cclrd{s}} \tau}
\end{equation*}
%
The last step uses the fact that 
$\cclrd{s} \,\aseq\, \alift{\cunit} = \alift{\cclrd{s} \,\cseq\, \cunit} = \alift{\cclrd{s}}$ 
arising from the monoid $(\C, \cseq, \cunit)$ of the scalar coeffect structure.

This highlights another difference between coeffects and effects, as $\eta$-equality does
not hold for many notions of effect. For example, in a language with output
effects, $e = (\textsf{print \texttt{"hi"}}; (\lambda x . x))$ has different effects to 
its $\eta$-converted form $\lambda x . e x$ because the immediate effects of $e$ are
hidden by the purity of $\lambda$-abstraction. In the coeffect calculus, the (abs) rule
allows immediate contextual requirements of $e$ to ``float outside'' of the enclosing
$\lambda$. Furthermore, the free monoid nature of $\cons$ in structural systems allows 
the exact immediate requirements of $\lambda x . e x$ to match those of $e$.

\subsection{Flat coeffect systems}

The equational theory for flat coeffect systems is somewhat similar to effect systems
where (co)effects are not linked to individual variables. In effectful languages,
substituting an effectful computation for $y$ in $\lambda x.y$ changes the latent effect 
associated with the function.

Similarly, for some of the flat coeffect systems, substituting a context-dependent computation
for $y$ in $\lambda x.y$ adds latent context requirements to the function type. However, this is not
the case for \emph{all} flat coeffect systems -- for example, call-by-name reduction preserves
types and coeffects for the implicit parameters system (which makes it a suitable model for Haskell).
For other systems, we first briefly consider call-by-value reduction.

\paragraph{What is a value?}
The notion of \emph{value} in coeffect systems differs from the usual syntactic understanding.
As discussed earlier, a function $(\lambda x.e)$ is not necessarily a value in coeffect calculi, 
because it may not delay all context requirements of $e$. Thus we say that $e$ is a value if it
has no immediate context requirements.

\begin{definition} 
An expression $e$ is a \emph{value}, written as $\textit{val}(e)$ if $\coctx{\Gamma}{ \aclrd{\mathsf{VAL}} } \vdash e : \tau$ 
where $\aclrd{\mathsf{VAL}} : \C^{\slift{\Gamma}}$ is a coeffect indexed by the shape of $\Gamma$ that
always returns $\cunit$. That is $\aclrd{\mathsf{VAL}} = \lambda \sclrd{n} . \cunit$.
\end{definition}

\paragraph{Call-by-value.}
In call-by-value, a pure \emph{value} is substituted for a variable. 
The right-hand side of an application is evaluated to a value before the
$\beta$-reduction, but the discharging of coeffects prior to substitution is
different for each concrete coeffect system.

Recall that a flat coeffect system consists of coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
together with a binary operation $\cclrd{\wedge}$ on $\C$. The coeffect algebra is then defined
as $(\cclrd{\wedge}, \cpar, \czero)$ where $\cclrd{\wedge}$ and $\cpar$ represent splitting and
merging of context-requirements, respectively.

\begin{lemma}[Call-by-value substitution]
\label{thm:flat-subst-cbv}
In a flat coeffect calculus with coeffect scalars $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\aclrd{\mathsf{VAL}}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\end{lemma}
\begin{proof}
By induction over the type derivation, using the fact that both $x$ and $e_s$ are annotated
with $\cunit$.
\end{proof}

\noindent
Lemma~\ref{thm:flat-subst-cbv} holds for all flat coeffect systems, but it is weak. To use it, 
the operational semantics must provide a way of partially evaluating a term 
with requirements $\coctx{\Gamma}{\cclrd{r}}$ to a value.
Assuming $\reduce_{\ident{cbv}}$ is call-by-value reduction using the above definition of value:

\begin{theorem}[Call-by-value reduction]
In a flat coeffect system, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \reduce_{\ident{cbv}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:flat-subst-cbv}, using the flat coeffect system requirement
$(\cclrd{r \wedge s}) \,\cleq\, (\cclrd{r}\,\cpar\,\cclrd{s})$ to prove $\beta$-equality.
\end{proof}

\paragraph{Call-by-name.}
The call-by-name strategy reduces a term $(\lambda x. e_1)~e_2$ where both sub-expressions 
may have contextual requirements. Assume that $\coctx{\Gamma}{\cclrd{s}} \vdash e_2 : \tau_2$ and
$\coctx{\Gamma_1, x :\tau_1}{\cclrd{r}} \vdash e_1 : \tau_1$.

% TODO
%
% The context requirements of the 
% reduced term $\subst{e_1}{x}{e_2}$ should be the same as the context requirements of 
% $\kvd{let}~x=e_2~\kvd{in}~e_1$. Thus the context requirements of the result in the substitution 
% lemma should be $r \fmax (s \fseq r)$.

We call a flat coeffect algebra \emph{top-pointed} if $\cunit$ is the greatest (top) coeffect scalar
$\C$ and \emph{bottom-pointed} if it is the smallest (bottom) coeffect scalar.
Liveness analysis is an example of top-pointed coeffects
as variables are annotated with $\ident{L}$ and $\ident{D} \leq \ident{L}$.

\begin{lemma}[Top-pointed substitution]
\label{thm:cbn-substitution-top}
In a top-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
Using sub-coeffecting ($\cclrd{s} \,\cleq\, \cunit$) and Lemma~\ref{thm:flat-subst-cbv}.
\end{proof}
%
\noindent
As variables are annotated with the top element $\cunit$, we can substitute a term $e_s$ for any variable and 
use sub-coeffecting to get the original typing (because $\cclrd{s} \,\cleq\, \cunit$).

In a bottom pointed coeffect system, substituting $e$ for $x$ increases the context 
requirements. However, if the system satisfies the strong condition that $\cclrd{\wedge} = \cseq = \cpar$ then
the context requirements arising from the substitution can be associated with the context
$\Gamma$. As a result, substitution does not break soundness as in effect systems. The 
requirement $\cclrd{\wedge} = \cseq = \cpar$ holds for our implicit parameters example 
(all three operators are set union) and allows the following substitution lemma:

\begin{lemma}[Bottom-pointed substitution]
\label{thm:cbn-substitution-bot}
In a bottom-pointed flat coeffect calculus with $(\C, \cseq, \cpar, \cunit, \czero, \cleq)$ 
and the $\cclrd{\wedge}$ operator where $\cclrd{\wedge} = \cseq = \cpar$ is idempotent and commutative:
\[
\begin{array}{l}
\coctx{\Gamma}{\cclrd{s}} \vdash e_s : \sigma \;\; \wedge \;\; \coctx{\Gamma_1,  x : \sigma, \Gamma_2}{ \cclrd{r}  } \vdash e_r : \tau\\
\quad \Rightarrow \;\; \coctx{\Gamma_1,\Gamma,\Gamma_2}{ \cclrd{r} \,\cseq\, \cclrd{s} } \vdash \subst{e_r}{x}{e_s} : \tau
\end{array}
\]
\vspace{-1.5em}
\end{lemma}
\begin{proof}
By induction over $\vdash$, using the idempotent, commutative monoid structure to keep $\cclrd{s}$ with
the free-variable context.
\end{proof}
\noindent
The structural system is precise enough to keep the coeffects associated with a concrete variable. The flat
variant described here is flexible enough to let us always re-associate new context requirements with the 
free-variable context.

The two substitution lemmas show that the call-by-name evaluation strategy can be used for
certain coeffect calculi, including liveness and implicit parameters.
Assuming $\reduce_{\ident{cbn}}$ is the standard call-by-name
reduction, the following theorem holds:

\begin{theorem}[Call-by-name reduction]
\raggedright
In a coeffect system that satisfies the conditions for Lemma~\ref{thm:cbn-substitution-top} or
Lemma~\ref{thm:cbn-substitution-bot}, if $\coctx{\Gamma}{\cclrd{r}} \vdash e : \tau$ 
and $e \rightarrow_{\ident{cbn}} e'$ then $\coctx{\Gamma}{\cclrd{r}} \vdash e' : \tau$.
\end{theorem}
\begin{proof}
A direct consequence of Lemma~\ref{thm:cbn-substitution-top} or Lemma~\ref{thm:cbn-substitution-bot}.
\end{proof}

>>>>

@@

\section{Semi-lattice formulation}
\label{sec:unified-semilattice}

\subsection{Semi-lattice formulation}
\subsection{Type inference algorithm}

@@

\newcommand{\tystmt}[2]{ (\text{\footnotesize{#1}})~~{#2} }
\newcommand{\ctxtrans}[3]{ #2 \rightsquigarrow #1, #3 }
\newcommand{\ctxtransnl}[3]{ \begin{array}{l} #2 \rightsquigarrow\\ #1, #3 \end{array}}

\newcommand{\SH}{ {\textcolor{sclr}{\mathcal{S}}} }
\newcommand{\sempty}{ {\textcolor{sclr}{ \hat{0}}} }
\newcommand{\sunit}{ {\textcolor{sclr}{\hat{1}}} }
\newcommand{\stimes}{ {\textcolor{sclr}{\diamond}} }
\newcommand{\sflat}{\star}

\newcommand{\amerge}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{-90}{$\ltimes$}} } }
\newcommand{\asplit}{ \textcolor{aclr}{ \mathbin{\rotatebox[origin=c]{90}{$\ltimes$}} } }
\newcommand{\azero}{ \textcolor{aclr}{\bot} }
\newcommand{\aunit}{ \alift{\cunit} } %\textcolor{aclr}{\textnormal{U}} }
\newcommand{\aweak}{ \alift{\czero} } % \textcolor{aclr}{\textnormal{I}} }

\newcommand{\fcons}{ \times } 
\newcommand{\fempty}{\epsilon}

\section{The unified coeffect calculus}

@@

\noindent
The operation $\cseq$ must form a monoid with $\cunit$ to guarantee an underlying 
category in the semantics. 
It models sequential composition with variable access
($\cunit$) as the identity. The other element ($\czero$) is used for
variables that are not accessed. The operation $\cpar$ combines coeffects
for contexts used in multiple places (contraction). 
The notation is inspired by the bounded reuse example, which uses coeffect scalar structure
$(\cclrd{\mathbb{N}}, \ast, +, 1,0, \cclrd{\leq})$, but be aware that $\cseq$ and $\cpar$ do
not always mean $\ast$ and $+$.

The context annotations can be viewed as \emph{containers} of scalar coeffects.
For structural coeffects, the container is a vector, while for flat coeffects, it is
a trivial singleton container. We take inspiration from the work of Abbott \emph{et al.}~\cite{types-containers}
which describes containers in terms of \emph{shapes} and a set of \emph{positions} in each shape.

@@

@@

\[
\tystmt{weak}
  { \quad\quad 
    \ctxtrans{ \coctx{\Gamma}{\aclrd{R}} }{ \coctx{\Gamma, x:\tau}{\aclrd{R}\; \asplit\; \aweak } }{\emptyset}
    \quad\quad\quad }
\]
\[ 
\tystmt{exch}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau, y:\sigma, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q}} }
               { \quad \coctx{\Gamma_1, y:\sigma, x:\tau, \Gamma_2}{\aclrd{R} \,\asplit\, \alift{\cclrd{t}} \,\asplit\, \alift{\cclrd{s}} \,\asplit\, \aclrd{Q}} }{\emptyset} }
\]
\[
\tystmt{contr}
  { \ctxtransnl{ \coctx{\Gamma_1, y:\tau, z:\tau, \Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \alift{\cclrd{t}} \,\amerge\, \aclrd{Q} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s}\, \cpar\, \cclrd{t}} \,\asplit\, \aclrd{Q}} }{[y,z \mapsto x]}}
\]
\[
\tystmt{sub}
  { \ctxtransnl{ \coctx{\Gamma_1, x:\tau,\Gamma_2}{\aclrd{R} \,\amerge\, \alift{\cclrd{s}} \,\amerge\, \aclrd{T} } }
               { \quad \coctx{\Gamma_1, x:\tau, \Gamma_2}{\aclrd{R}\, \asplit\, \alift{\cclrd{s'}} \,\asplit\, \aclrd{T}} }{\emptyset} }
\quad(\cclrd{s} \;\cleq\; \cclrd{s'} )                 
\]
\caption{The general coeffect calculus}
\vspace{-1em}
\label{fig:unified-coeffects}
\end{figure}

@@

\subsection{Structural coeffects}
\label{sec:unified-structural}

@@

\subsection{Flat coeffects}
\label{sec:unified-flat}

@@
